Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-04-13 mpaper/paper.tex
Major refactoring

Imagine using a machine learning (ML) algorithm to teach a robot how to move
around people so that it learns to predict where people are going and adjust its
path accordingly. The ML algorithm would use data about various possible
situations. But do we have enough data to ensure reasonably optimal behaviour?
Perhaps the robot behaves well in most situations, but fails in less common
scenarios. Can the ML model itself describe its weaknesses so that we could
ensure it is exposed to sufficiently many uncommon or difficult situations?

This learning problem
\cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} as well as
many others have benefited from an approach called \emph{inverse reinforcement
  learning} (IRL), also known as apprenticeship learning and inverse optimal
control. IRL proposes a way to learn behaviour from demonstrations that
typically come from human actions. More formally, the IRL problem asks us to
find a reward function for a Markov decision process (MDP), where demonstrations
are encoded as sets of state-action pairs.

IRL is an important problem because adjusting the reward function by hand is
often unwieldy, since human behaviour often depends on many factors in
complicated ways
\cite{DBLP:conf/icml/PieterN04}. Moreover,
learning the reward function rather than the policy itself makes the model more
transferable to new environments---a minor change in the environment can
reorganise the whole policy but only have a local effect in the reward structure
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
IRL has been used to teach helicopters how to perform tricks
\cite{DBLP:conf/nips/AbbeelCQN06}, predict taxi destinations
\cite{DBLP:conf/huc/ZiebartMDB08}, and make driving safer and more efficient by
predicting pedestrian movement \cite{DBLP:conf/iros/ZiebartRGMPBHDS09} and the
driver's intentions \cite{DBLP:conf/aaai/VogelRGR12}.

Most IRL models in the literature make a convenient yet unjustified
assumption that the reward function can be expressed as a linear combination of
features
\cite{DBLP:conf/icml/PieterN04,DBLP:conf/icml/NgR00,ziebart2008maximum}.
This assumption severely restricts what reward structures can be modelled (i.e.,
the properties of the function that represents rewards across states). Out of
the non-linear models proposed to date, none can answer the questions posed in
the first paragraph. More specifically, without knowing the ground truth for the
learning problem, we have no reliable way to know whether the learned behaviour
is optimal and which parts of the model could benefit from more data. Quite
often, the models assume that rewards have no variance
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}.

@@

% TODO: could raise intractable posterior as the main reason, but would have to
% rephrase the whole section
No fully Bayesian nonlinear IRL model has been proposed yet. We will uncover one
reason for this when we set up our IRL model for VI in Section~\ref{sec:model}.
Applying VI to this model in a theoretically sound way (without introducing
unjustified simplifying assumptions) requires us to use Lebesgue's dominated
convergence theorem in order to derive the gradient of the expectation of the
MDP value function---we provide a complete proof for this (and some other
results) in Section~\ref{sec:proof}. Finally, in Section~\ref{sec:evaluation} we
demonstrate the model's correctness and properties experimentally using several
example scenarios.

@@

For any matrix $\mathbf{A}$, we use either $A_{i,j}$ or
$[\mathbf{A}]_{i,j}$ to denote the element of $\mathbf{A}$ in row $i$ and column
$j$. We use $\tr(\mathbf{A})$ to denote its \emph{trace} and $\adj(\mathbf{A})$
for its \emph{adjugate} (or \emph{classical adjoint}). For any vector
$\mathbf{x}$, we write $\mathbb{R}_d[\mathbf{x}]$ to denote a vector space of
polynomials with degree at most $d$, where variables are elements of
$\mathbf{x}$, and coefficients are in $\mathbb{R}$.

@@

\section{Background} % 0.5-1 page
% TODO: edit this: we can probably extend this (and the introduction)

The IRL problem itself was originally proposed by Russell in 1998
\cite{DBLP:conf/colt/Russell98}. Most of the early approaches had the
aforementioned reward linearity assumption. One of the first papers on the
subject by Ng and Russell \cite{DBLP:conf/icml/NgR00} introduced several linear
programming algorithms and identified an important issue: there are typically
many reward functions that can explain the data equally well. This problem was
solved by Ziebart et al. \cite{ziebart2008maximum} with the introduction of IRL
based on the principles of maximum causal entropy in a linearly-solvable MDP.

Levine et al. \cite{DBLP:conf/nips/LevinePK11} were the first to lift the
linearity assumption without imposing additional restrictions on the problem.
They do, however, model rewards as having no variance---our work removes this
restriction without any compromises.

@@

We argue that developing an IRL model with a variational approximation of the
full posterior distribution has eluded previous research in the area for two
reasons. First, it is not directly relevant to the goal of producing accurate
point estimates of rewards. However, we described several key benefits in the
previous section, and we postulate how this work could affect the wider fields
of reinforcement learning and artificial intelligence in
Section~\ref{sec:conclusions}. Second, as it will become apparent in
Sections~\ref{sec:model} and \ref{sec:proof}, optimising such a VI model
involves estimates of gradients, establishing correctness of which requires a
significant amount of work.

@@

A major development in approximating complex distributions came in the form of
\emph{normalizing flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. A sequence of such
functions can transform a simple initial distribution (e.g., uniform or
Gaussian) into an arbitrarily complex one.
%A major challenge in applying this
%work to IRL is related to variational parameters being used to compute the MDP
%value function, i.e., how can we take the derivative of the value function with
%respect to such a variational parameter? Alternatively, perhaps one can
%construct a different model that would make this question moot.

Another approach to flexibility in modelling could come from considering
different GP kernels. For instance, Wilson and Adams \cite{pmlr-v28-wilson13}
show how all \emph{stationary} (i.e., invariant to translations) kernels can be
generated (or at least approximated) from a mixture of Gaussians in their
spectral representation using Bochner's theorem
\cite{bochner1959lectures,DBLP:journals/technometrics/Woodard00}. It looks
promising to combine these kernels with the variational Fourier features
approach by Hensman et al. \cite{DBLP:journals/jmlr/HensmanDS17} that leverages
the same spectral representations for efficient VI.

While these and many other advancements in VI promise great improvements, there
are hidden challenges when it comes to applying them to VI. Many suggested
improvements implicitly impose restrictions on the model that are unlikely to be
satisfied in the context of IRL, e.g., the evidence of the model being a
Gaussian. Nevertheless, the work outlined here seems adaptable to the IRL
setting and can make variational approximations both more accurate and faster.

@@

In this section, we introduce definitions and mathematical details relevant to
the problem. We begin by formally defining the problem.

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, \mathbf{r} \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state
  $s'$ after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor; and $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$ is the reward
  vector.
\end{definition}

While it is more common to represent rewards as a function $r : \mathcal{S} \to
\mathbb{R}$, the vector notation is generally more appropriate in our situation.
However, we switch between the two notations as needed.

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of \emph{demonstrations} $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs that represents optimal (or
  near-optimal) actions, find the reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

\begin{example}
  Figure~\ref{fig:example} shows a possible setup for IRL. In this case,
  $\mathcal{S} = \{ s_1, s_2, s_3, s_4 \}$, $\mathcal{A} = \{ l, r, u, d \}$,
  \begin{align*}
    \mathcal{T}(s, a, s') &= \begin{cases}
      1 & \text{if } (s, a, s') = (s_1, r, s_2) \\
      1 & \text{if } (s, a, s') = (s_1, d, s_4) \\
      1 & \text{if } (s, a, s') = (s_2, l, s_1) \\
      1 & \text{if } (s, a, s') = (s_2, d, s_3) \\
      1 & \text{if } (s, a, s') = (s_3, l, s_4) \\
      1 & \text{if } (s, a, s') = (s_3, u, s_2) \\
      1 & \text{if } (s, a, s') = (s_4, r, s_3) \\
      1 & \text{if } (s, a, s') = (s_4, u, s_1) \\
      0 & \text{otherwise,}
    \end{cases} \\
      \mathbf{X} &= \begin{blockarray}{ccc}
        & f_1 & f_2 \\
        \begin{block}{c[cc]}
          s_1 & 1 & 2 \\
          s_2 & 2 & 2 \\
          s_3 & 2 & 1 \\
          s_4 & 1 & 1 \\
        \end{block}
      \end{blockarray},
  \end{align*}
  and $\gamma \in [0, 1)$ can be assigned any value. The distribution of example
  state-action pairs into separate demonstrations is irrelevant to our model
  (and possible all other models), so we can keep each pair in its own set as
  follows:
  \[
    \mathcal{D} = \{ \{(s_1, d)\}, \{(s_2, l)\}, \{(s_4, u)\}, \{(s_4, u)\} \}.
  \]

  A solution to this problem is then an assignment of rewards to states, e.g.,
  \[
    \mathbf{r} = \begin{bmatrix}
      2 \\
      0 \\
      -1 \\
      1
    \end{bmatrix} \quad \text{or} \quad r(s) = \begin{cases}
      2 & \text{if } s = s_1 \\
      0 & \text{if } s = s_2 \\
      -1 & \text{if } s = s_3 \\
      1 & \text{if } s = s_4.
    \end{cases}
  \]
\end{example}

@@

    \draw [<->] (-1.3, 0.8) node (yaxis) [above] {$f_2$}
    |- (3.5, -3.6) node (xaxis) [right] {$f_1$};
    \draw[shift={(0,-3.6)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$1$};
    \draw[shift={(2.65,-3.6)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$2$};
    \draw[shift={(-1.3,-2.6)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$1$};
    \draw[shift={(-1.3,0)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$2$};
  \end{tikzpicture}
  \caption{An example MDP with deterministic actions. States are represented by
    grey circles, actions by directed labelled edges between them. The numbers
    in parentheses next to some actions' names denote how many times that
    state-action pair appears in $\mathcal{D}$. If that number is zero, the edge
    is dashed. Finally, the axes assign two features to each state.}
  \label{fig:example}
\end{figure}

The set of demonstrations $\mathcal{D}$ often comes from observed human
behaviour and constitutes the most significant part of the data. The optimal
(deterministic) \emph{policy} $\pi : \mathcal{S} \to \mathcal{A}$ (i.e., a
choice of actions for each state that maximises reward over time) is usually
constructed by defining a \emph{value (utility) function} $\V : \mathcal{S} \to
\mathbb{R}$ that measures how good a state is based on the reward $\mathbf{r}$
as well as the structure of the MDP. One can then find $\V$ by applying the
Bellman backup operator until convergence to every $s \in \mathcal{S}$ (the
technique is known as \emph{value iteration}) \cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

However, we follow previous work on GP IRL
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}, and use a
\emph{linearly solvable} (or \emph{maximum causal entropy}) MDP with a
stochastic policy that defines probability distributions over actions (instead
of suggesting a single action for each state) \cite{ziebart2008maximum}. This
type of MDP can be solved by applying the `soft' version of the operator
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}:
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
With this model, we can express the likelihood as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{align*}
  p(\mathcal{D} \mid \mathbf{r}) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} \mid s_{i,t}) \\
                                 &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right),
\end{align*}
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s').
\]

A significant part of learning is about being able to generalise to previously
unseen input states (otherwise the problem becomes knowledge representation
instead of learning). For this reason, we represent each state by a
$d$-dimensional feature vector. This way, similar states are located in close
proximity to one another, and the ML model can predict rewards for new states
based on their similarity to training data.

In this paper, we model rewards as a function from feature space to
$\mathbb{R}$ using a \emph{Gaussian process}. A GP represents an infinite
collection of random variables, any finite number of which has a joint Gaussian
distribution \cite{DBLP:books/lib/RasmussenW06}. A GP is defined by its mean
function (which is always $\mathbf{x} \mapsto 0$ in our case) and covariance
function, which we denote by $k$. \emph{Covariance functions} (also known as
\emph{kernels}) take two state feature vectors as input and quantify how similar
the two states are, in a sense that we would expect high covariance scores to be
associated with similar rewards.

@@

We can then use VI in order to approximate
$p(\mathbf{u},~\mathbf{r}~\mid~\mathcal{D})$. This is done by minimising the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation. KL divergence (asymmetrically) measures the
difference between two probability distributions. Let $\approximation$ denote
our approximation. The KL divergence is then defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$, so we can remove it from our optimisation objective. The
negation of what remains is known as the \emph{evidence lower bound} (ELBO) and
is defined as \cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}_{\approximation}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters to be defined in Section~\ref{sec:model}.
Note that hereafter we drop the subscript notation, as all expected values will
be with respect to $\approximation$.

@@

In order to have a fully functional model, we still need to make several key
design decisions. In this section, we describe the covariance function used to
define the GP, consider what parameters should be used to optimise $\mathcal{L}$
and how they should be initialised, and in Section~\ref{sec:elbo}, we derive our
final expression for $\mathcal{L}$.

@@

Ideally, we would like to model $\bm\lambda$ with an approximating distribution.
However, due to how the prior probability density function (PDF) for
$\mathbf{u}$ involves $\Kuu^{-1}$, and $\mathcal{L}$ has an expected value that
cannot be eliminated (see Section~\ref{sec:elbo}), we are unable to show that
such an $\mathcal{L}$ is well-defined. More generally, we pose the following
problem, which, to the best of our knowledge, is currently open:
\begin{problem}
  Let $\mathbf{A}$ be a $n \times n$ matrix of coefficients, $X$ be a random
  variable, and $\mathbf{M}$ be an $n \times n$ matrix such that $M_{i,j} = f(X,
  A_{i,j})$, where $f$ is an arbitrary function. Under what circumstances does
  $\mathbb{E}[\mathbf{M}^{-1}]$ exist?
\end{problem}
While there are some obvious examples where the expected value exists (e.g.,
$f(X, A_{i,j}) = A_{i,j}X$ for an invertible $\mathbf{A}$ and many distributions
of $X$), it would be particularly interesting to know whether the answer is
`always'. A proof of such a result would allow us to model $\bm\lambda$ instead
of treating it as a variational parameter, and would help to guard against
overfitting. For now, $\bm\lambda$ will have to be treated as a variational
parameter.

>>>>

Imagine using a machine learning (ML) algorithm to teach a robot how to move
around people so that it learns to predict where people are going and adjust its
path accordingly. The ML algorithm would learn from data about proper actions in
a range of different situations. But do we have enough data to ensure reasonably
optimal behaviour? Perhaps the robot behaves well in most situations, but fails
in less common scenarios. Can the ML model itself describe its weaknesses so
that we could ensure it is exposed to sufficiently many uncommon or difficult
situations?

This learning problem \cite{DBLP:journals/ijrr/KretzschmarSSB16} as well as
many others have benefited from an approach called \emph{inverse reinforcement
  learning} (IRL), also known as apprenticeship learning and inverse optimal
control. IRL proposes a way to learn behaviour from demonstrations that
typically come from human actions. More formally, the IRL problem asks us to
find a reward function for a Markov decision process (MDP), where demonstrations
are encoded as sets of state-action pairs, showing observed actions in a number
of states.

IRL is an important problem because adjusting the reward function by hand is
often unwieldy, since human behaviour often depends on many factors in
complicated ways
\cite{DBLP:conf/icml/PieterN04}. Moreover,
learning the reward function rather than the policy itself makes the model more
transferable to new environments: a minor change in the environment can
reorganise the whole policy but only have a local effect in the reward structure
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
IRL has been used to teach helicopters how to perform tricks
\cite{DBLP:conf/nips/AbbeelCQN06}, predict taxi destinations
\cite{DBLP:conf/huc/ZiebartMDB08}, and make driving safer and more efficient by
predicting pedestrian movement \cite{DBLP:conf/iros/ZiebartRGMPBHDS09} and the
driver's intentions \cite{DBLP:conf/aaai/VogelRGR12}.

Most IRL models in the literature make a convenient yet unjustified
assumption that the reward function can be expressed as a linear combination of
features
\cite{DBLP:conf/icml/PieterN04,DBLP:conf/icml/NgR00,ziebart2008maximum}.
This assumption severely restricts what reward structures can be modelled. Out
of the non-linear models proposed to date, none can answer the questions posed
in the first paragraph. More specifically, without knowing the ground truth for
the learning problem, we have no reliable way to know whether the learned
behaviour is optimal and which parts of the model could benefit from more data.
Quite often, the models assume that rewards have no variance
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}.

@@

No fully Bayesian nonlinear IRL model has been proposed yet. We will uncover one
reason for this when we set up our IRL model for VI in Section~\ref{sec:model}.
Applying VI to this model in a theoretically sound way (and without introducing
unjustified simplifying assumptions) requires us to use Lebesgue's dominated
convergence theorem in order to derive the gradient of the expectation of the
MDP value function---we provide a complete proof for this (and some other
results) in Section~\ref{sec:proof}. Finally, in Section~\ref{sec:evaluation} we
demonstrate the model's correctness and properties experimentally using several
example scenarios.

@@

For any matrix $\mathbf{A}$, we write either $A_{i,j}$ or
$[\mathbf{A}]_{i,j}$ to denote the element of $\mathbf{A}$ in row $i$ and column
$j$. We use $\tr(\mathbf{A})$ to denote its \emph{trace} and $\adj(\mathbf{A})$
for its \emph{adjugate} (or \emph{classical adjoint}). For any vector
$\mathbf{x}$, we write $\mathbb{R}_d[\mathbf{x}]$ to denote a vector space of
polynomials with degree at most $d$, where variables are elements of
$\mathbf{x}$, and coefficients are in $\mathbb{R}$.

@@

\section{Background} % 0.5-1 page

The IRL problem itself was originally proposed by Russell in 1998
\cite{DBLP:conf/colt/Russell98}. Most of the early approaches had the
aforementioned reward linearity assumption. One of the first papers on the
subject by Ng and Russell \cite{DBLP:conf/icml/NgR00} introduced several linear
programming algorithms and identified an important issue: there are typically
many reward functions that can explain the data equally well. This problem was
solved by Ziebart et al. \cite{ziebart2008maximum} with the introduction of IRL
based on the principle of maximum causal entropy in a linearly-solvable MDP.

Levine et al. \cite{DBLP:conf/nips/LevinePK11} were the first to lift the
linearity assumption without imposing additional restrictions on the problem by
modelling rewards with a Gaussian process (GP). They do, however, assume that
rewards have no variance---our work removes this restriction without any
compromises.

@@

We argue that developing an IRL model with a variational approximation of the
full posterior distribution has eluded previous research in the area for two
reasons. First, it is not directly relevant to the goal of producing accurate
point estimates of rewards. However, we described several key benefits of
variance prediction in the previous section, and we postulate how this work
could affect the wider fields of reinforcement learning and artificial
intelligence in Section~\ref{sec:conclusions}. Second, as it will become
apparent in Sections~\ref{sec:model} and \ref{sec:proof}, optimising such a VI
model involves estimating gradients, and establishing correctness of these
estimates requires a significant amount of theoretical groundwork.

@@

A major development in approximating complex distributions came in the form of
\emph{normalizing flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. A sequence of such
functions can transform a simple initial distribution (e.g., uniform or
Gaussian) into an arbitrarily complex one.

Another approach to flexibility in modelling can be achieved by considering
different GP kernels, i.e., functions that model similarity between pairs of
points. For instance, Wilson and Adams \cite{pmlr-v28-wilson13} show how all
\emph{stationary} (i.e., invariant to translations) kernels can be generated (or
at least approximated) from a mixture of Gaussians in their spectral
representation using Bochner's theorem \cite{bochner1959lectures}. It looks
promising to combine these kernels with the variational Fourier features
approach by Hensman et al. \cite{DBLP:journals/jmlr/HensmanDS17} that leverages
the same spectral representations for efficient VI. This algorithm, however, is
currently restricted to Mat\'ern kernels, which could be restrictive,
considering how IRL models tend to use kernels that can adjust the importance of
each feature \cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.

While these and many other advancements in VI promise great improvements, there
are hidden challenges when it comes to applying them to IRL. Many suggested
improvements implicitly impose restrictions on the model that are unlikely to be
satisfied in the context of IRL, e.g., the evidence of the model being a
Gaussian. Nevertheless, the work mentioned here seems adaptable to the IRL
setting and can make variational approximations both more accurate and faster.

@@

In this section, we describe the problem in more detail, introducing the
notation and definitions that will be used throughout the paper.

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, \mathbf{r} \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T}
  \colon \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a
  function defined so that $\mathcal{T}(s, a, s')$ is the probability of moving
  to state $s'$ after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the
  discount factor; and $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$ is the reward
  vector.
\end{definition}

While it is more common to represent rewards as a function $r \colon \mathcal{S}
\to \mathbb{R}$, the vector notation is generally more appropriate in our
situation. However, we switch between the two notations as needed.

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of \emph{demonstrations} $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs that represents optimal (or
  near-optimal) actions, find a reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

\begin{example}
  Figure~\ref{fig:example} shows a possible setup for IRL. In this case,
  $\mathcal{S} = \{ s_1, s_2, s_3, s_4 \}$, $\mathcal{A} = \{ l, r, u, d \}$,
  \begin{align*}
    \mathcal{T}(s, a, s') &= \begin{cases}
      1 & \text{if } (s, a, s') = (s_1, r, s_2) \\
      1 & \text{if } (s, a, s') = (s_1, d, s_4) \\
      1 & \text{if } (s, a, s') = (s_2, l, s_1) \\
      1 & \text{if } (s, a, s') = (s_2, d, s_3) \\
      1 & \text{if } (s, a, s') = (s_3, l, s_4) \\
      1 & \text{if } (s, a, s') = (s_3, u, s_2) \\
      1 & \text{if } (s, a, s') = (s_4, r, s_3) \\
      1 & \text{if } (s, a, s') = (s_4, u, s_1) \\
      0 & \text{otherwise,}
    \end{cases} \\
      \mathbf{X} &= \begin{blockarray}{ccc}
        & f_1 & f_2 \\
        \begin{block}{c[cc]}
          s_1 & 1 & 2 \\
          s_2 & 2 & 2 \\
          s_3 & 2 & 1 \\
          s_4 & 1 & 1 \\
        \end{block}
      \end{blockarray},
  \end{align*}
  and $\gamma \in [0, 1)$ can be assigned any value. The distribution of example
  state-action pairs into separate demonstrations is irrelevant to our model
  (and presumably all other models) and is only a relic of the practical
  motivation behind the problem's initial definition. Hence, we can keep each
  pair in its own set as follows:
  \[
    \mathcal{D} = \{ \{(s_1, d)\}, \{(s_2, l)\}, \{(s_4, u)\}, \{(s_4, u)\} \}.
  \]

  A solution to this problem is then an assignment of rewards to states, e.g.,
  \[
    \mathbf{r} = \begin{bmatrix}
      2 \\
      0 \\
      -1 \\
      1
    \end{bmatrix} \quad \text{or} \quad r(s) = \begin{cases}
      2 & \text{if } s = s_1 \\
      0 & \text{if } s = s_2 \\
      -1 & \text{if } s = s_3 \\
      1 & \text{if } s = s_4
    \end{cases}.
  \]
\end{example}

@@

    \draw [<->] (-1.3, 0.8) node (yaxis) [above] {$f_2$}
    |- (3.5, -3.6) node (xaxis) [right] {$f_1$};
    \draw[shift={(0,-3.6)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$1$};
    \draw[shift={(2.65,-3.6)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$2$};
    \draw[shift={(-1.3,-2.6)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$1$};
    \draw[shift={(-1.3,0)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$2$};
  \end{tikzpicture}
  \caption{An example MDP with deterministic actions. States are represented by
    grey circles, actions by directed labelled edges between them. The numbers
    in parentheses next to some actions' names denote how many times that
    state-action pair appears in $\mathcal{D}$. If that number is zero, the edge
    is dashed. Finally, the axes are used to assign two features to each state.}
  \label{fig:example}
\end{figure}

The optimal (deterministic) \emph{policy} $\pi \colon \mathcal{S} \to
\mathcal{A}$ (i.e., a choice of actions for each state that maximises reward
over time) is usually constructed by defining a \emph{value (utility) function}
$\V \colon \mathcal{S} \to \mathbb{R}$ that measures how good a state is based
on the reward $\mathbf{r}$ as well as the structure of the MDP. One can then
find $\V$ by applying the Bellman backup operator until convergence to all
states $s \in \mathcal{S}$ (this algorithm is known as \emph{value iteration})
\cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

However, we follow previous work on IRL with GPs
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11} and use a
\emph{linearly solvable} (or \emph{maximum causal entropy}) MDP with a
stochastic policy that defines probability distributions over actions instead
of suggesting a single action for each state \cite{ziebart2008maximum}. This
type of MDP can be solved by applying the `soft' version of the operator
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}:
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
With this model, we can express the likelihood as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{align*}
  p(\mathcal{D} \mid \mathbf{r}) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} \mid s_{i,t}) \\
                                 &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right),
\end{align*}
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s').
\]

A significant part of learning is about being able to generalise to previously
unseen input states (otherwise the problem becomes knowledge representation
instead of learning). For this reason, we represent each state by a
$d$-dimensional feature vector. This way, similar states are represented by
similar features, and the ML model can predict rewards for new states based on
their similarity to training data.

In this paper, we model rewards as a function from feature space to
$\mathbb{R}$ using a \emph{Gaussian process}. A GP represents an infinite
collection of random variables from a joint (infinitely-dimensional) probability
distribution, any finite number of which has a Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. A GP is defined by its mean function (which
is always $\mathbf{x} \mapsto 0$ in our case) and covariance function, which we
denote by $k$. \emph{Covariance functions} (also known as \emph{kernels}) take
two state feature vectors as input and quantify how similar the two states are,
in a sense that we would expect high covariance scores to be associated with
similar rewards.

@@

We can then use VI in order to approximate
$p(\mathbf{u},~\mathbf{r}~\mid~\mathcal{D})$. This is done by minimising the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation. KL divergence (asymmetrically) measures the
difference between two probability distributions. Let $\approximation$ denote
our approximation. The KL divergence is then defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$, so we can remove it from our optimisation objective. The
negation of what remains is known as the \emph{evidence lower bound} (ELBO) and
is defined as \cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}_{\approximation}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters that we define in
Section~\ref{sec:model}. Note that hereafter we drop the subscript notation, as
all expected values will be with respect to $\approximation$.

@@

In order to have a fully functional model, we still need to make several key
design decisions. In this section, we describe the covariance function used to
define the GP, consider what parameters should be used to optimise $\mathcal{L}$
and how they should be initialised, and in Section~\ref{sec:elbo} we derive our
final expression for $\mathcal{L}$.

@@

Ideally, we would like to model $\bm\lambda$ with an approximating distribution.
However, due to how the prior probability density function (PDF) for
$\mathbf{u}$ involves $\Kuu^{-1}$, and $\mathcal{L}$ has an expected value that
cannot be eliminated (see Section~\ref{sec:elbo}), we are unable to show that
such an $\mathcal{L}$ is well-defined. More generally, we pose the following
problem, which, to the best of our knowledge, is currently open:
\begin{problem}
  Let $\mathbf{A}$ be a $n \times n$ matrix of coefficients, $X$ be a random
  variable, and $\mathbf{M}$ be an $n \times n$ matrix such that $M_{i,j} = f(X,
  A_{i,j})$, where $f$ is an arbitrary function. Under what circumstances does
  $\mathbb{E}[\mathbf{M}^{-1}]$ exist?
\end{problem}
While there are some obvious examples where the expected value exists (e.g.,
$f(X, A_{i,j}) = A_{i,j}X$ for an invertible $\mathbf{A}$ and many distributions
of $X$), it would be particularly interesting to know whether the answer is
`always'. A proof of such a result would allow us to model $\bm\lambda$ instead
of treating it as a variational parameter, which would make the model fully
Bayesian. For now, $\bm\lambda$ will have to be treated as a variational
parameter.

It remains to decide on the approximating distribution for $\mathbf{u}$ and
$\mathbf{r}$. As is commonly done when applying VI to GPs
\cite{DBLP:conf/nips/ChengB17}, we set
\begin{equation} \label{eq:approximation}
  \approximation = q(\mathbf{u}) q(\mathbf{r} \mid \mathbf{u}),
\end{equation}
where $q(\mathbf{r} \mid \mathbf{u}) = p(\mathbf{r} \mid \mathbf{u})$ and
$q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \bm\mu, \bm\Sigma)$.
